Predicting Success of Clinical Interviews via Probabilistic Modeling of Patient-Provider Communication Sequences

### 2017-03-08 paper/AMIA/main.tex
Intro updated by Alex

\title{Predicting Success of Clinical Interviews via Sequential Analysis of Patient-Provider Interactions}

@@

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered sequences of past events to make predictions related to the observed sequence of events arises in many data analytics scenarios in healthcare informatics. In this paper, we focus on the analysis of patient-provider communication sequences in the context of clinical interviews and propose the methods for identifying the most effective communication sequences for triggering a particular type of patient response and dynamically predicting this type of response based on observed interactions. Such analysis is necessary to facilitate the development of effective interventions to address many public health conditions, such as childhood obesity. Our proposed probabilistic method for predicting a patient behavioral response in communication sequences can be utilized with both Markov chains or Hidden Markov Models. It achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using first and second order Markov chain and Hidden Markov Model, respectively. Our results indicate that individual short length patient-provider communication sequences are highly indicative of the progression and trajectory of clinical interviews and can be used to predict the overall success of motivational interviews. Our results have broad implications for qualitative public health research by providing a formal mechanism to facilitate the development of more effective behavioral interventions.}

@@

Sequence classification has a broad range of real word applications from genomics and health informatics to finance and abnormality detection. In genomics research, sequence classification is widely used to detect the function of new proteins \cite{deshpande2002evaluation}. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}. Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware detection \cite{drew2017polymorphic}.

In this paper, we direct our focus towards patient-provider communication sequences in the context of clinical interviews. In particular, we propose a method to identify the most effective communication sequences for eliciting a specific type of patient response and . Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment is an important part of clinical research aimed at designing effective interventions for many behavioral conditions and disorders. In this study, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents (teens) and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. Therefore, there is a need for informatics-based methods to facilitate the development of effective interventions for childhood obesity. One approach to designing an effective intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview.

Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes. In this study, probabilistic models are employed for the analysis of sequence data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language. To the best of our knowledge, there have been very limited studies on sequence classification in a motivational interview such as patient-provider dialogue. A sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to find the relationship between physician and patient behaviors associated with patients' exhibiting cues. Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns. Although hidden Markov model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification. In addition, a k-order Markov model is applied to genomic research \cite{yakhnenko2005discriminatively} to classify protein and text sequence data. Therefore, we utilize Markov chain and hidden Markov model for the classification of utterance sequences.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental data sets for this work were constructed based on the transcripts of 37 motivational interview sessions, which include a total of 11,353 segmented and annotated utterances. Since clinical researchers tried to increase the desire and ability of an adolescent to change their current behavior to target behavior, we used sequences from an adolescent session and excluded sequences from caregiver session in the MI transcript. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 different behavior codes, which are grouped into the youth, caregiver, and counselor code groups. These utterances are then scanned from first to last and divided into successful and unsuccessful sequences when a commitment or change talk is found. A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one data set that includes all sequences of positive change talk or commitment language (880 sequences) and the other data set which includes utterances ended with negative change talk or commitment language (192 sequences). Alternate sequences are the possible combination of a normal sequence where two consecutive codes can not come from the same speaker. For example, there are 6 possible alternate sequences for the following sequences: (C1, A1, A2, C2, C3, C4, A3), where adolescent and counselor codes are represented by A and C, respectively. Therefore, (C1, A1, C2, A3), (C1, A1, C3, A3), (C1, A1, C4, A3), (C1, A2, C2, A3), (C1, A2, C3, A3) and (C1, A2, C4, A3) were used as alternate sequences for the above sequence. A total of 18753 alternate sequences has been observed which further partitioned into two subsets of 18139 successful and 614 unsuccessful sequences. In our experiment, normal and alternate sequences are employed with two subsets (successful and unsuccessful) to train Markov chain and HMM. Dataset statistics and a fragment of an adolescent session transcript are presented in Table~\ref{tab:data_dist} and Table~\ref{tab:anno_examp}. \\

@@

\subsection*{\textit{Sequence classification techniques}}
Generally, a sequence is an ordered list of events. In our study, an event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.  Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$. Here, L is fixed and known in advances. In our experiment, known labels are ``successful'' and ``unsuccessful'' motivational interview.

>>>>

\title{Predicting Success of Clinical Interviews via Probabilistic Modeling of Patient-Provider Communication Sequences}

@@

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered sequences of past events to make predictions related to the observed sequence of events arises in many data analytics scenarios in healthcare informatics. In this paper, we focus on automating the analysis of patient-provider communication sequences in the context of clinical interviews and propose the methods for identifying the most effective communication exchanges to elicit a particular type of patient response and dynamically predicting this type of response in an observed patient-provider communication. Our proposed method for predicting the likelihood of desired patient behavioral response in communication sequences can be used in conjunction with either Markov chain or Hidden Markov Model and can significantly decrease the effort  required to develop effective interventions to address many public health conditions, such as childhood obesity. It achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using first and second order Markov chain and Hidden Markov Model, respectively,  to model communication sequences. Our results indicate that individual short length patient-provider communication sequences are highly indicative of the overall progression and future trajectory of a clinical interview and can be used to predict its overall success. Our results have broad implications for qualitative public health research by providing a formal mechanism to facilitate the development of effective behavioral interventions.}

@@

Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection. In genomics research, sequence classification is widely used to classify protein and text sequence data \cite{yakhnenko2005discriminatively} and detect the function of new proteins \cite{deshpande2002evaluation}. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}. Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware \cite{drew2017polymorphic}.

In this paper, we direct our focus towards patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to designing an effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview. However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual annotation and examination of transcripts is a very time-consuming process, development of new obesity interventions can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions.

While our previous work \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing 115 patient and provider behavior codes \cite{carcone2013provider}, in the work, we propose probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview.

While sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to establish the relationship between physician and patient behaviors associated with patients' exhibiting cues, there have been only a handful of previous studies on modeling annotated sequences of patient-provider communication exchanges and predicting a desired patient behavior in the context of clinical interviews.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 37 motivational interviews, which include a total of 11,353 segmented and annotated utterances. Each motivational interview consists of counselor-adolescent and counselor-caregiver sessions. Since the ultimate goal of a motivational interview is to increase the desire and ability of an adolescent for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 distinct behavior codes, which are grouped into the youth, caregiver, and counselor code groups. These utterances are then scanned from first to last and divided into successful and unsuccessful communication sequences. A successful communication sequence results in positive change talk or commitment languages, while an unsuccessful sequence results in or change talk is found. A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one data set that includes all sequences of positive change talk or commitment language (880 sequences) and the other data set which includes utterances ended with negative change talk or commitment language (192 sequences). Alternate sequences are the possible combination of a normal sequence where two consecutive codes can not come from the same speaker. For example, there are 6 possible alternate sequences for the following sequences: (C1, A1, A2, C2, C3, C4, A3), where adolescent and counselor codes are represented by A and C, respectively. Therefore, (C1, A1, C2, A3), (C1, A1, C3, A3), (C1, A1, C4, A3), (C1, A2, C2, A3), (C1, A2, C3, A3) and (C1, A2, C4, A3) were used as alternate sequences for the above sequence. A total of 18753 alternate sequences has been observed which further partitioned into two subsets of 18139 successful and 614 unsuccessful sequences. In our experiment, normal and alternate sequences are employed with two subsets (successful and unsuccessful) to train Markov chain and HMM. Dataset statistics and a fragment of an adolescent session transcript are presented in Table~\ref{tab:data_dist} and Table~\ref{tab:anno_examp}. \\

@@

\subsection*{\textit{Sequence classification techniques}}

 Although hidden Markov model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification.
model patient-provider interactions using first- and second-order Markov Chains and Hidden Markov Models (HMM)
popular probabilistic methods for modeling discrete observation sequences with finite vocabulary,
Generally, a sequence is an ordered list of events. In our study, an event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.  Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$. Here, L is fixed and known in advances. In our experiment, known labels are ``successful'' and ``unsuccessful'' motivational interview.
